Towards typed repositories of proofs

### 2010-07-05 talks/MIPS2010/slides.tex
slides mips2010: Partie motivations bien avancées
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@137 403e655e-980a-47bd-8c60-1ee5348bb6b7

\usepackage{ucs}
\usepackage{mathpartir}
\usepackage{amsfonts,amsmath,amscd}
\usepackage{stmaryrd}
\usepackage[utf8x]{inputenc}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{dot2texi}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}

@@

\AtBeginSection[]
{\begin{frame}<beamer>{Outline}
    \tableofcontents[currentsection]
  \end{frame}
}

@@

\begin{frame}
  \titlepage
  \mode<article>{
    \newcommand\url\texttt
    \maketitle}
\end{frame}
\section*{Introduction}

@@

\begin{frame}{A paradoxical situation}
  Most time is spent \emph{editing}, not \emph{writing} proofs.

  Workflow of formal mathematics is largely inspired by software
  development:
  \begin{itemize}
  \item Text-based
  \item Separate compilation (\textsf{make})
  \item Versioning (\textsf{cvs}, \textsf{svn}, \textsf{diff}s\ldots)
  \end{itemize}
\end{frame}

Let me start with two simple observations relating the activity of
both the working mathematician and the programmer. A first, obvious
fact is that both spend more time \emph{editing} than \emph{writing}:
their workflow is highly non-linear, involving experiments, fixes,
backtrack on previous modifications\ldots, eventually validated or
invalidated by some criteria like the absence of bugs or the validity
of a proof.

The second observation is that since the advent of modern proof
assistants, large bodies of mathematics have been formalized and the
way people managed these developments is very similar to the one used
in software development: we write proofs in text buffers, which are
linear, and launch proof-checking when we are done. To avoid
re-checking the whole development, we split it into mutually dependent
files and re-check only those who have changed. To keep track of old
versions, we use line-based version control systems that have no
awareness of the actual content of the files. The separation into
multiple files then serves two different purpose, and the fact that
they coincide is pure contingency: the purpose of splitting a
development into logical units, and the purpose of easing the
non-linear process of proof editing. Also, recording the actual text
of the proofs for versioning is often non-satisfactory: on one side it
contains mostly non-relevant informations (line breaks, spaces\ldots),
on the other the relevant ones, the validity of the proof, the
dependency between objects, is absent.

The paradox in this situation is that these very tools can be used to
reason about formal languages and especially proof languages. Couldn't
we adapt the methods developped to reason about a formal language to
the proof language of proof assistants itself? Could we replace this
legacy toolchain and make it language-aware?

\begin{frame}{Expectations}

@@

\begin{frame}{\mode<beamer>{Outline}}
  \tableofcontents
\end{frame}

\section{}

\subsection{Practical applications}

>>>>

\usepackage{ucs}
\usepackage{mathpartir}
\usepackage{amsfonts,amsmath,amscd}
\usepackage{stmaryrd}
\usepackage[utf8x]{inputenc}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{setspace}

@@

% \AtBeginSection[]
% {\begin{frame}<beamer>{Outline}
%     \tableofcontents[currentsection]
%   \end{frame}
% }

@@

\begin{frame}
  \titlepage
  \mode<article>{
    \newcommand\url\texttt
    \maketitle}
\end{frame}

\section{Motivation}

@@

\begin{frame}{A paradoxical situation}
  \begin{block}{Observation 1}
    Most of a programmer/mathematician's time is spent \emph{editing}, not \emph{writing}
  \end{block}
  \begin{block}{Observation 2}
    Workflow of formal mathematics was largely inspired by software
    development:
    \begin{itemize}
    \item File-based scripts
    \item Separate compilation (\textsf{make})
    \item Text-based versioning (\textsf{cvs}, \textsf{svn},
      \textsf{diff}s\ldots)
    \end{itemize}
  \end{block}
  \begin{block}{Postulate}
    \it Current administrative toolchain should be adapted to proof languages
  \end{block}
  {\Large }
\end{frame}

Let me start with some simple observations relating the activity of
both the working mathematician and the programmer. A first, obvious
fact is that both spend more time \emph{editing} than \emph{writing}:
their workflow is highly non-linear, involving experiments, fixes,
backtrack on previous modifications\ldots, eventually validated or
invalidated by some criteria like the absence of bugs or the validity
of a proof.

The second observation is that since the advent of modern proof
assistants, large bodies of mathematics have been formalized and the
way people managed these developments is very similar to the one used
in software development: we write proofs in text buffers, which are
linear, and launch proof-checking when we are done. To avoid
re-checking the whole development, we split it into inter-dependent
files and re-check only those who have changed. To keep track of old
versions, we use line-based version control systems that have no
awareness of the actual content of the files. The separation into
multiple files serves two different purpose, and the fact that they
coincide is pure contingency: the purpose of splitting a development
into logical units, and the purpose of easing the non-linear process
of proof editing. Also, recording the actual text of the proofs for
versioning is often non-satisfactory: on one side it contains mostly
non-relevant informations (line breaks, spaces\ldots), on the other
the relevant ones, the validity of the proof, the dependency between
objects, is absent.

The paradox in this situation is that these very tools can be used to
reason about formal languages and especially proof languages. Couldn't
we adapt the methods developped to reason about formal languages to
the language of proof assistants itself? Could we replace this legacy
toolchain and make it language-aware?

\begin{frame}{Motivations}
  Current toolset is not adapted to proof edition:
  \begin{itemize}
  \item ((edit; compile)*; commit)* loop does not scale to proofs
  \item Concept freeze inhibits the discovery process
  \item Any change in a file requires full recompilation of
    dependencies
  \item No room for alternate definitions (linearity of scripts)
  \item Textual scripts \textsf{diff}s do not reflect the semantics
  \item Not even the syntax
  \end{itemize}
  \vspace{2em}
  {\tiny \ldots\ Maybe it wasn't adapted to software development}
\end{frame}

More precisely, here are some problems that arise with the traditional
tools when editing proofs. Whereas the usual edit, compile, commit
loop may be a satisfactory approximation when developing software, the
additional demands of proof-checking makes it unsatisfactory as the
development grows: often the time of recompilation is prohibitive.

But it is not only a matter of time: once a concept is proof-checked,
it is frozen and any future modification of it will require to
re-check a large part of the development, often much more than what is
actually required. We believe that this facts inhibits the use of
proof assistants as tools to help the mathematician in the discovery
process, and not only to formally transcribe paper proofs.

The same way, the linearity of proof scripts does not allow any
flexibility in the alternate definition of concepts.

Finally of course, the way we store mathematical developments, that is
as textual files or textual diffs does not reflect neither their syntax
nor their semantics, and as I will try to show, there is a lot to gain
in changing this base representation.

In fact, some will say that this model of interaction wasn't even
adapted to programming in the first place.

\begin{frame}{The impact of changes}
  These problems were addressed by a number of (ad-hoc) fixes:
  \begin{itemize}
  \item File-based separate compilation
  \item Interface/implementation
  \item Interaction loop with global undo
  \end{itemize}
  \vspace{2em}
  \begin{center} \Large
    \begin{spacing}{1.3}
      \textit{Need for a refined notion of dependency to analyze the
        impact of changes}
    \end{spacing}
  \end{center}
\end{frame}

We can then see a number of extra-logical features of modern proof
assistants as ways to cope with these problems. Starting from a
hypothetical ``dark age'' of proof assistants where one would write
its complete development in one unique file and let the program check
it when done, were added:
\begin{itemize}
\item The ability to split it into files organized in a dependency
  graph as discussed earlier, which is a coarse approximation of
  actual dependencies amongst objects
\item The ability to separate interfaces (statements) from
  implementations (proofs), and recheck dependencies only if
  implementations changed.
\item Most proof assistants base their interactivity on a mechanism
  for backtracking their entire states globally, as a way to make
  changes to already checked parts.
\end{itemize}

\begin{frame}{Example: \textsf{coqide}}

@@

\begin{frame}{Methodology}
  \begin{center}
    \only<1-2>{\fbox{\parbox{20em}{\only<2>\alert{version control}\\[1em]%
        \fbox{\parbox{18em}{script files\\[1em]%
            \fbox{\parbox{16em}{parsing\\[1em]%
                \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
    \only<3>{\fbox{\parbox{20em}{script files\\[1em]%
        \fbox{\parbox{18em}{\alert{version control}\\[1em]%
            \fbox{\parbox{16em}{parsing\\[1em]%
                \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
    \only<4->{\fbox{\parbox{20em}{script files\\[1em]%
        \fbox{\parbox{18em}{parsing\\[1em]%
            \fbox{\parbox{16em}{\alert{version control \only<5->{+ proof-checking}}\\[1em]%
                \uncover<4>{\fbox{\parbox{14em}{proof-checking\\}}}}}}}}}}%
  \end{center}
  \uncover<6>{
    \begin{itemize}
    \item AST representation
    \item Typing annotations
    \item Incremental type-checking
    \end{itemize}
  }
\end{frame}
